Specific leaf mass explains difference in biomass partitioning between woody plant functional types

### 2014-10-21 manuscript/draftpaper.Rnw
Text improvements, new appendix on predicting basal D from DBH, code reorganized a bit. Figure 1 has PFT averages for MAP and MAT.

It has been reported numerous times that gymnosperms (with higher SLM) have higher leaf mass fraction than angiosperms (with lower SLM) \cite{enquist_global_2002, cheng_relationships_2007, mccarthy_consistency_2007, poorter_biomass_2012, sardans_tree_2013}, although a recent study reported no difference (\cite{wolf_allometric_2010}). No study to date has been able to analyze differences between deciduous and evergreen angiosperms, or link differences in partitioning between PFTs to other plant traits such as SLM.

@@

We consider here the relationship between leaf biomass and area with stem diameter (an important allometric relationship, Enquist others, and related to the pipe model if the ratio sapwood / heartwood is relatively constant, an important measure of hydraulic supply), the aboveground leaf mass fraction (partioning of biomass into leaf vs. woody pools), and the leaf area ratio (a measure of the photosynthetic area per unit woody biomass). We expect that $M_F$ and $A_F$ are approximately proportional to the square of stem diameter, which would remove size-related shifts in allometry that often confound interpretations. For each of these variables, we ask whether 1) there are differences between major plant functional types after removing size-related variation, 2) there are differences due to climate, compared between major biomes, or as a function of the mean annual temperature (as recently reported for whole stand biomass, \cite{reich_temperature_2014}).

@@

 The ratio of aboveground biomass to basal stem area did not differ by PFT (Fig.~\ref{fig:figureSI5}).

 However we found very large differences between PFTs in the ratio of leaf mass to basal stem area (Fig.~\ref{fig:figure3}), but the ratio leaf area to basal stem area converged between PFTs.

 We then expressed the average $M_F/A_s$ as a function of the specific leaf mass (SLM = $M_F / A_F$), a measure of the leaf construction cost to the plant. We found that, across PFTs, $M_F/A_S$ was approximately proportional to the average specific leaf mass for the PFT. As a result of this approximate proportionality, the ratio of leaf area to basal stem area ($A_F/A_S$) was not different between PFTs (Fig.~\ref{fig:figure4}) (see also Fig.~\ref{fig:figureSI6} for raw data).

@@

 Unlike \cite{reich_temperature_2014}, we found only very weak effects of mean annual temperature the scaling of leaf biomass with woody biomass (Fig.~\ref{fig:figureSI9}). The size-invariant relationship was only significant for deciduous angiosperms, but it was a very weak correlation that explained very little additional variation. These results were confirmed with a mixed-effects model of MF vs. MS, with MAT and PFT as covariates (see Table XXX). We found no significant effects of mean annual precipitation of leaf-woody biomass relationships (Table XXX).

 We found coordination between an important leaf trait and whole-plant biomass partitioning. This has implications for global vegetation models.

@@

 A recent study compared allocation routines in a number of leading ecosystem models \cite{de2014does}, and recommended to constrain allocation by observed biomass fractions instead of using constant allocation fractions, which is common practice.

 As a first approximation, which is useful for large scale DGVMs, models should aim to keep leaf area to stem cross-sectional area constant, and vary biomass allocation accordingly. Specific leaf area then gives the ratio of leaf biomass to stem biomass.

 We found no appreciable effects of climate on anything, when analyzed in addition to PFT. This means that climate-related differences are captured by a shift in PFT (see also Fig.~\ref{figure1}, showing PFTs occupy different parts of the climate space). Our results contrast with those of \cite{reich_temperature_2014} but they studied whole-stand biomass, not individual plant biomass. So perhaps this means the climate effects are on stand density not individual plant effects. No effect of climate is good news for models that parameterize the globe by PFT, as this seems sufficient, based on our dataset. Are there experimental data that show temperature effects on allocation? What about water availability?

@@

 BAAD database (Falster et al., in review), including a total of \Sexpr{nrow(baad$data)} observations, but here we use only data that include leaf mass ($M_F$), leaf area ($A_F$), aboveground woody biomass ($M_W$), total root biomass ($M_R$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (ca. 1.3m) ($A_{Sbh}$).  For each analysis, we therefore used different subsets of the data (far fewer observations are available for root biomass, in particular), sample sizes are included in the caption of each figure (\emph{to be added}). We further excluded deciduous gymnosperms (since very few data were available), and glasshouse and common garden studies, including only field grown woody plants (including plantations).

 We focus here only on aboveground biomass because we found no PFT differences in root-shoot scaling (Fig.~\ref{fig:figureSI6}), and approximately isometric scaling (sma slope = xxx) (consistent with other reports, \cite{hui_near_2014, cairns_root_1997}).

@@

 We can further decompose aboveground biomass ($M_T$) into a size-related metric (basal stem area, $A_S$), and the ratio of $M_T$ to $A_S$, which is a function of wood density and plant woody volume to basal area.

We test for PFT differences to both these components of biomass partioning. We were unable to further split $A_S/M_W$ because wood density data were available only for a subset of data. Chave has shown that etc. etc.

 We considered three plant functional types : evergreen angiosperms, evergreen gymnosperms, and decidous angiosperms. All locations were further separated into boreal (including sub-boreal), temperate, and tropical biomes (decided by contributing author, but checked with MAT and latitude).

 We did not attempt here to include SLA from another database (such as TRY), instead only used SLA directly estimated for the harvested plants.

 Stem area was measured either at breast height (ca. 1.3m) or at the base of the plant (normally for small plants). For the subset of the data where both were measured, we estimated basal stem area (Asba) from breast height area (Asbh) by fitting XXX equation. Results were similar when we used either measure of stem area (not shown).

 We used a generalized additive model (GAM) to visualize the relationships, but these were not used for formal inference.

 We used linear mixed-effects models (with species within study as the random effect) to test for significance and to calculate confidence intervals for PFT-specific averages. Least-square means for LMF and LAR were calculated with the 'lsmeans' package.

 All variables were log-transformed prior to analysis, to stabilize variance.

 Variance partitioning with marginal (fixed effects) and conditional (fixed and random effects) $R^2$ for linear mixed-effects models as reported by \cite{nakagawa_general_2013}.

 For conifers, leaf area was converted to half-total surface area, which is most appropriate in terms of light interception  \cite{lang_application_1991, chen_defining_1992}, for comparison to flat leaves.

@@

\begin{figure}[h!]
    \centering
    \includegraphics{Figure1_MAPMAT_baad_vs_worldclim.pdf}
    \caption{Global coverage of the climate space by the dataset. Hexagons are terrestrial observations extracted from Worldclim (colored by number of 0.5\degree~cells). Note that Worldclim observations include land areas where woody vegetation does not occur (e.g., desert, tundra). The deciduous gymnosperms were not included in any analysis in this paper, but are shown in the figure.  }
    \label{fig:figure1}
\end{figure}

@@

<<tablepreload>>=
load("tables/Table_counts.RData")
printit <- function(x, caption){
  print(xtable(x, align="rlll|l",
               caption=caption
               ), hline.after=c(0,3), caption.placement="top")

@@

<<results='asis'>>=
printit(tab_mlfastba, caption="Number of observations used in the analysis of $M_F/A_S$ and $A_F/A_S$. Individual plants (number of species). DA = Deciduous Angiosperm, EA = Evergreen Angiosperm, EG = Evergreen Gymnosperm.")
@

% <<results='asis'>>=
% printit(tab_alfastba, caption="Number of observations used in the analysis of Af/As. Individual plants (number of species). DA = Deciduous Angiosperm, EA = Evergreen Angiosperm, EG = Evergreen Gymnosperm.")
% @
%

<<results='asis'>>=
printit(tab_lmf, caption="Number of observations used in the analysis of leaf mass fraction (LMF) and leaf area ratio (LAR). Individual plants (number of species). DA = Deciduous Angiosperm, EA = Evergreen Angiosperm, EG = Evergreen Gymnosperm.")
@

% 
% <<results='asis'>>=
% printit(tab_lar, caption="Number of observations used in the analysis of leaf area ratio (LAR). Individual plants (number of species). DA = Deciduous Angiosperm, EA = Evergreen Angiosperm, EG = Evergreen Gymnosperm.")
% @

>>>>

It has been reported numerous times that gymnosperms (with higher SLM) have higher leaf mass fraction than angiosperms (with lower SLM) \cite{enquist_global_2002, cheng_relationships_2007, mccarthy_consistency_2007, poorter_biomass_2012, sardans_tree_2013}, although a recent study reported no difference (\cite{wolf_allometric_2010}). No study to date has been able to further tease apart PFT differences, or test the potentially confounding effects of climate.

@@

We found no PFT differences in root-shoot scaling (Fig.~\ref{fig:figureSI6}), and approximately isometric scaling (sma slope = xxx), which is consistent with other reports, \cite{hui_near_2014, cairns_root_1997}). Because of this simple scaling between above- and below-ground biomass pools, we further focus on biomass partioning between leaves and aboveground woody pools.

The main objective was to test whether PFT-specific differences exist in the ratio of leaf mass to total above-ground biomass (the 'leaf mass fraction', LMF), and the leaf area to aboveground biomass (the 'leaf area ratio', LAR). Because PFT and climate are confounded to some extent (Fig.~\ref{figure1}), we also test whether climate has additional effects on biomass partitioning, or whether these effects are captured by PFT alone.

@@

We can further decompose LMF into two components representing different aspects of plant function and structure. The ratio of leaf biomass ($M_F$) to stem basal area ($A_S$) relates to hydraulic supply, and the ratio of total aboveground biomass ($M_T$) and $A_S$ is a function of plant shape (stem taper, plant height at given stem diameter), and wood density \cite{chave_tree_2005}.

We found that ratio of aboveground biomass to basal stem area ($M_T/A_S$) did not differ by PFT (Fig.~\ref{fig:figureSI5}). However we found very large differences between PFTs in the ratio $M_F/A_S$(Fig.~\ref{fig:figure3}).

 We then expressed the average $M_F/A_S$ as a function of the specific leaf mass (SLM = $M_F / A_F$), a measure of the leaf construction cost to the plant. We found that, across PFTs, $M_F/A_S$ was approximately proportional to the average specific leaf mass for the PFT. As a result of this approximate proportionality, the ratio of leaf area to basal stem area ($A_F/A_S$) was not different between PFTs (Fig.~\ref{fig:figure4}) (see also Fig.~\ref{fig:figureSI6} for raw data).

@@

 Unlike \cite{reich_temperature_2014}, we found only very weak effects of mean annual temperature the scaling of leaf biomass with woody biomass (Fig.~\ref{fig:figureSI9}). However, the variance explained by MAT was very small, compared to the large difference between PFTs. These results were confirmed with a mixed-effects model of MF vs. MS, with MAT and PFT as covariates (see Table~\ref{table:matlmeanova}). We found no significant effects of mean annual precipitation of leaf-woody biomass relationships (Table~\ref{table:maplmeanova}).

In summary, we found coordination between an important leaf trait and whole-plant biomass partitioning. This has implications for global vegetation models.

@@

A recent study compared allocation routines in a number of leading ecosystem models \cite{de2014does}, and recommended to constrain allocation by observed biomass fractions instead of using constant allocation fractions, which is common practice. As a first approximation, which is useful for large scale DGVMs, models should aim to keep leaf area to stem cross-sectional area constant, and vary biomass allocation accordingly. Specific leaf area then gives the ratio of leaf biomass to stem biomass.

We found no effects of MAP on biomass partitioning, and only very weak effects of MAT on LMF, when analyzed in addition to PFT. This means that climate-related differences are captured by a shift in PFT (see also Fig.~\ref{figure1}, showing PFTs occupy different parts of the climate space). Our results contrast with those of \cite{reich_temperature_2014} but they studied whole-stand biomass, not individual plant biomass. So perhaps this means the climate effects are on stand density not individual plant effects. No effect of climate is good news for models that parameterize the globe by PFT, as this seems sufficient, based on our dataset.

@@

We used the recently compiled Biomass And Allometry Database (BAAD) (Falster et al., in review), including a total of \Sexpr{nrow(baad$data)} observations, but here we use only data that include leaf mass ($M_F$), leaf area ($A_F$), aboveground woody biomass ($M_W$), total root biomass ($M_R$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (ca. 1.3m) ($A_{Sbh}$).  For each analysis, we therefore used different subsets of the data (far fewer observations are available for root biomass, in particular). Sample sizes by PFT are summarized in Tables~\ref{table:count1}-\ref{table:count4} We exluded glasshouse and common garden studies, including only field grown woody plants (including plantations). We considered three plant functional types : evergreen angiosperms, evergreen gymnosperms, and decidous angiosperms. We excluded  deciduous gymnosperms because few data were available. All locations were further separated into boreal (including sub-boreal), temperate, and tropical biomes (decided by contributing author, but checked with MAT and latitude).

@@

We were unable to further decompose $A_S/M_W$ into wood volume and wood density components (cf. \cite{chave_tree_2005}) because wood density data were available only for a small subset of data.
We did not include SLM estimates from another database (such as TRY), instead only used SLA directly estimated for the harvested plants (typically for a subsample of leaves). For conifers, leaf area was converted to half-total surface area, which is most appropriate in terms of light interception  \cite{lang_application_1991, chen_defining_1992}, for comparison to flat leaves.

Stem cross-sectional area was measured either at breast height (ca. 1.3m) and/or at the base of the plant. For the subset of the data where both were measured, we estimated basal stem area ($A_S$) from breast height stem area ($A_{Sbh}$) using non-linear regression (see~SIXXX). The main results were similar when we repeated it for data where either $A_S$ or $A_{Sbh}$ were directly measured (not shown).

We used a generalized additive model (GAM) to visualize the relationships, but these were not used for formal inference. Instead we used linear mixed-effects models (with species within study as the random effect) to test for significance and to calculate confidence intervals for PFT-specific averages. Least-square means for LMF and LAR were calculated with the 'lsmeans' package in R. All variables were log-transformed prior to analysis. Variance partitioning with marginal (fixed effects) and conditional (fixed and random effects) $R^2$ for linear mixed-effects models as reported by \cite{nakagawa_general_2013}.

@@

\begin{figure}[h!]
    \centering
    \includegraphics{Figure1_MAPMAT_baad_vs_worldclim.pdf}
    \caption{Global coverage of the climate space by the dataset. Hexagons are terrestrial observations extracted from Worldclim (colored by number of 0.5\degree~cells, cut off at 6000mm MAP). Note that Worldclim observations include land areas where woody vegetation does not occur (e.g., desert, tundra). Triangles are averages for the three PFTs across all contributed studies. }
    \label{fig:figure1}
\end{figure}

@@

<<tablepreload1>>=
load("tables/Table_counts.RData")
printit <- function(x, caption, label=NULL){
  print(xtable(x, align="rlll|l",
               caption=caption, label=label
               ), hline.after=c(0,3), caption.placement="top")

@@

<<results='asis'>>=
printit(tab_mlfastba, 
        label="table:count1",
        caption="Number of observations used in the analysis of $M_F/A_S$ and $A_F/A_S$. Individual plants (number of species). DA = Deciduous Angiosperm, EA = Evergreen Angiosperm, EG = Evergreen Gymnosperm.")
@

<<results='asis'>>=
printit(tab_alfastba, 
        label="table:count2",
        caption="Number of observations used in the analysis of Af/As. Individual plants (number of species). DA = Deciduous Angiosperm, EA = Evergreen Angiosperm, EG = Evergreen Gymnosperm.")
@

<<results='asis'>>=
printit(tab_lmf, 
        label="table:count3",
        caption="Number of observations used in the analysis of leaf mass fraction (LMF) and leaf area ratio (LAR). Individual plants (number of species). DA = Deciduous Angiosperm, EA = Evergreen Angiosperm, EG = Evergreen Gymnosperm.")
@

<<results='asis'>>=
printit(tab_lar, 
        label="table:count4",
        caption="Number of observations used in the analysis of leaf area ratio (LAR). Individual plants (number of species). DA = Deciduous Angiosperm, EA = Evergreen Angiosperm, EG = Evergreen Gymnosperm.")
@

<<tablepreload2>>=
library(car)
load("tables/Fits_lme_mlfmst_MAPMAT.RData")
printanova <- function(x, caption, label=NULL){
  rownames(x) <- gsub("log10(m.st)", "MS", rownames(x), fixed=TRUE) 
  rownames(x) <- gsub("pft", "PFT", rownames(x)) 
  print(xtable(x, caption=caption, label=label),
                                        caption.placement="top")
}
@

<<results='asis'>>=
printanova(Anova(mlfmst_lme1), 
           label="table:matlmeanova",
           caption="Type-II Anova table for a linear mixed-effects model of $M_F$ as a function of $M_S$ (both log-transformed), PFT and mean annual temperature (MAT). Species by dataset was used as the random effect. Further analysis of the interactions revealed that MAT only significantly affected the relationship between $M_F$ and $M_S$ for the deciduous angiosperms (not shown).")
@

<<results='asis'>>=
printanova(Anova(mlfmst_lme2), 
           label="table:maplmeanova",
           caption="Type-II Anova table for a linear mixed-effects model of $M_F$ as a function of $M_S$ (both log-transformed), PFT and mean annual precipitation (MAT). Species by dataset was used as the random effect. We did not analyze MAT and MAP together in one model, as they were strongly correlated, making interpretation of the fitted model difficult.")
@

\clearpage
\section{Supporting Information Methods}

\subsection*{Predicting basal diameter}

<<>>=
load("tables/basalafit.RData")
@

Using the subset of data where basal stem diameter ($A_S$) and diameter at breast height ($A_{Sbh}$) were measured (a total of \Sexpr{length(residuals(basalafit))} observations covering the three major PFTs), we developed a non-linear model to predict $A_S$ when only $A_{Sbh}$ and plant height ($H$) were measured. We fit the following equation, using non-linear regression.

$$
D_S = D_{Sbh} H^a / (H - H_{bh})^a
$$

where $D_S$ is the basal stem diameter (m), $D_{Sbh}$ stem diameter at breast height, $H_{bh}$ the height at which $D_{Sbh}$ was measured (typically 1.3 or 1.34m), and $a$ was further expressed as a function of plant height:

$$
a = c_0H^c_1
$$

The estimated coefficients were $c_0$ = \Sexpr{coef(basalafit)[1]}, $c_1$ = \Sexpr{coef(basalafit)[2]}, root-mean square error = \Sexpr{summary(basalafit)$sigma}.
